\chapter*{References}
\addcontentsline{toc}{chapter}{References}
\markboth{References}{References}

The following sources are cited throughout this book using bracketed notation [S\#].

\section*{Academic Research}

\begin{description}[leftmargin=3em, labelwidth=2.5em, labelsep=0.5em]

\item[\textbf{[S1]}] Dell'Acqua, Fabrizio et al. ``Navigating the Jagged Technological Frontier: Field Experimental Evidence of the Effects of AI on Knowledge Worker Productivity and Quality.'' \textit{California Management Review}, 2024. Harvard Business School study on AI productivity effects.

\item[\textbf{[S12]}] Dell'Acqua, Fabrizio et al. ``Falling Asleep at the Wheel: Human/AI Collaboration in a Field Experiment on HR Recruiters.'' Harvard Business School Working Paper, 2024. Key finding: workers with AI completed 12\% more tasks, 25\% faster, with 40\% higher quality.

\item[\textbf{[S13]}] Brynjolfsson, Erik et al. ``Generative AI at Work.'' \textit{Science}, 2023. Research on the floor/ceiling effect of AI on productivity.

\item[\textbf{[S14]}] Brynjolfsson, Erik et al. ``The Effects of Generative AI on High Skilled Work.'' \textit{Quarterly Journal of Economics}, 2024. Finding: 14\% faster task completion, 13.8\% more queries resolved.

\item[\textbf{[S17]}] Dellermann, Dominik et al. ``The Future of Human-AI Collaboration: A Taxonomy of Design Knowledge for Hybrid Intelligence Systems.'' \textit{Hawaii International Conference on System Sciences}, 2019. Framework for co-evolutionary AI-human models.

\item[\textbf{[S18]}] Yao, Shunyu et al. ``ReAct: Synergizing Reasoning and Acting in Language Models.'' \textit{arXiv preprint}, 2022. Foundation for agentic reasoning frameworks.

\item[\textbf{[S21]}] Wake, Bill. ``INVEST in Good Stories, and SMART Tasks.'' XP123.com, 2003. Origin of the INVEST framework for user stories.

\item[\textbf{[S23]}] Lucassen, Garm et al. ``Improving Agile Requirements: The Quality User Story Framework and Tool.'' \textit{Requirements Engineering}, 2016. Research on systematic requirements quality.

\item[\textbf{[S25]}] Forsgren, Nicole et al. ``DORA State of DevOps Report.'' Google Cloud, 2018-2024. Research on elite team deployment patterns and organizational performance.

\item[\textbf{[S27]}] Forsgren, Nicole, Jez Humble, and Gene Kim. \textit{Accelerate: The Science of Lean Software and DevOps.} IT Revolution Press, 2018. Key finding: elite teams deploy 208x more frequently than low performers.

\item[\textbf{[S38]}] Turing, Alan. ``Computing Machinery and Intelligence.'' \textit{Mind}, 1950. Foundation of the Turing Test.

\item[\textbf{[S39]}] Wikipedia contributors. ``AI winter.'' \textit{Wikipedia, The Free Encyclopedia}. Historical account of AI funding contractions and company failures.

\item[\textbf{[S46]}] International Journal of Scientific Research and Publications (IJSRP). ``AI Ethics Frameworks: A Systematic Review.'' 2023. Research on accountability gaps in AI deployment.

\item[\textbf{[S50]}] Khojah, Ranim et al. ``Beyond the Hype: Developer Perspectives on AI-Powered Code Generation.'' \textit{IEEE Software}, 2024. Research on developer resistance to AI tools.

\item[\textbf{[S51]}] Gradle Inc. ``2023 State of Developer Productivity Report.'' Finding: perceived productivity doesn't correlate with actual output.

\item[\textbf{[S52]}] Cai, Zixuan et al. ``Understanding AI Leverage Skills in Software Development.'' \textit{Empirical Software Engineering}, 2024.

\item[\textbf{[S53]}] Edmondson, Amy. \textit{The Fearless Organization: Creating Psychological Safety in the Workplace for Learning, Innovation, and Growth.} Wiley, 2019.

\item[\textbf{[S54]}] International Journal of Scientific Research and Advancements (IJSRA). ``Generative AI in Product Management: A Systematic Review.'' 2024.

\item[\textbf{[S55]}] Russell, Stuart and Peter Norvig. \textit{Artificial Intelligence: A Modern Approach.} Chapter on ``The Dark Ages of AI.'' Pearson, 2020.

\end{description}

\section*{Industry Reports and Frameworks}

\begin{description}[leftmargin=3em, labelwidth=2.5em, labelsep=0.5em]

\item[\textbf{[S2]}] California Management Review. ``Building Competitive Advantage with GenAI.'' CMR Insights, 2024. Framework for sustainable AI-driven differentiation.

\item[\textbf{[S3]}] California Management Review. ``Three Organizational Traps in GenAI Adoption.'' CMR Insights, 2024. Analysis of premature scaling, capability overestimation, and governance gaps.

\item[\textbf{[S5]}] Deloitte. ``Tech Trends 2026: The AI Revolution.'' Deloitte Insights, 2025. Introduction of ``adaptive capacity'' as core organizational competency.

\item[\textbf{[S6]}] Deloitte. ``State of Generative AI in the Enterprise.'' Q3 2024. Key finding: 79\% of executives expect transformation within 3 years.

\item[\textbf{[S7]}] Deloitte. ``The State of Trust in AI.'' Deloitte AI Institute, 2024. Research on trust calibration in AI systems.

\item[\textbf{[S8]}] Deloitte. ``Ethics of AI: A Framework for Responsible AI.'' Deloitte AI Institute, 2023. Core obligations: transparency, accountability, fairness, privacy-by-design.

\item[\textbf{[S9]}] McKinsey \& Company. ``The Economic Potential of Generative AI.'' June 2023. Estimate: AI can accelerate time-to-market by 10-20\%.

\item[\textbf{[S10]}] McKinsey \& Company. ``The State of AI in 2024: Gen AI Adoption Spikes and Starts to Generate Value.'' McKinsey Global Survey, 2024. Analysis of workflow transformation patterns.

\item[\textbf{[S15]}] GitHub. ``Research: Quantifying GitHub Copilot's Impact on Developer Productivity and Happiness.'' GitHub Blog, 2022. Finding: developers completed tasks 55\% faster with Copilot.

\item[\textbf{[S30]}] National Institute of Standards and Technology (NIST). ``AI Risk Management Framework (AI RMF 1.0).'' U.S. Department of Commerce, 2023. Framework for managing AI risks including hallucinations.

\item[\textbf{[S31]}] McKinsey \& Company. ``The Product Management Profession in the Age of AI.'' 2024. Describes PMs becoming ``orchestrators of AI-powered workflows.''

\item[\textbf{[S36]}] Google. ``Rules of Machine Learning: Best Practices for ML Engineering.'' Google Developers, 2023. Canonical guide for ML product development.

\item[\textbf{[S37]}] Wikipedia contributors. ``ChatGPT.'' \textit{Wikipedia, The Free Encyclopedia}. Documentation of ChatGPT's adoption milestones.

\item[\textbf{[S47]}] Office of the Director of National Intelligence (ODNI). ``Artificial Intelligence Ethics Framework for the Intelligence Community.'' 2024. Emphasis on continuous monitoring over one-time certification.

\item[\textbf{[S48]}] KDnuggets. ``AI Governance Trends for 2024-2025.'' Analysis of emerging regulatory patterns: mandatory disclosure, algorithmic auditing, sector-specific compliance.

\end{description}

\section*{Legal and Regulatory Guidance}

\begin{description}[leftmargin=3em, labelwidth=2.5em, labelsep=0.5em]

\item[\textbf{[S44]}] Wikipedia contributors. ``Artificial intelligence and copyright.'' \textit{Wikipedia, The Free Encyclopedia}. Overview of the evolving legal landscape around AI and copyright.

\item[\textbf{[S45]}] U.S. Copyright Office. ``Copyright Registration Guidance: Works Containing Material Generated by Artificial Intelligence.'' Federal Register, March 2023. Guidance that AI-generated content generally lacks human authorship required for copyright.

\end{description}

\section*{Books and Practitioner Resources}

\begin{description}[leftmargin=3em, labelwidth=2.5em, labelsep=0.5em]

\item[\textbf{[S4]}] Autor, David and Simon Johnson. ``Seven Myths About AI and Productivity.'' \textit{MIT Sloan Management Review}, 2024.

\item[\textbf{[S11]}] IBM Institute for Business Value. ``CEO Guide to Generative AI.'' 2024. Analysis of compute costs and infrastructure requirements.

\item[\textbf{[S16]}] ACM Queue. ``AI-Assisted Code Generation: Implications for Software Development.'' 2024.

\item[\textbf{[S19]}] Mitchell, Eric et al. ``DetectGPT: Zero-Shot Machine-Generated Text Detection Using Probability Curvature.'' 2023.

\item[\textbf{[S20]}] ISO/IEC/IEEE 29148:2018. ``Systems and software engineering---Life cycle processes---Requirements engineering.''

\item[\textbf{[S22]}] Cohn, Mike. \textit{User Stories Applied: For Agile Software Development.} Addison-Wesley, 2004.

\item[\textbf{[S24]}] Fabbrini, Fabrizio et al. ``The QuARS Tool for the Quality Assessment of Requirements Specifications.'' 2001.

\end{description}

\vspace{2em}
\noindent\textit{Note: All URLs accessed December 2024--January 2026. Some sources may require institutional access.}
