\chapter*{Preface}
\addcontentsline{toc}{chapter}{Preface}

\epigraph{Train up a child in the way he should go: and when he is old, he will not depart from it.}{Proverbs 22:6}

I did not plan to write this book. I planned to figure out fatherhood quietly, privately, the way men are supposed to---absorbing information, making mistakes in silence, and eventually emerging on the other side as someone who ``knows what he's doing.''

That's not what happened.

What happened was chaos. What happened was fear dressed up as confidence. What happened was crying in the bathroom at 3 a.m. while my wife nursed our newborn and I wondered why I felt like a stranger in my own life. What happened was a slow, difficult, beautiful transformation that I didn't ask for but desperately needed.

This book is the manual I wish someone had handed me. Not a parenting guide full of developmental milestones and expert opinions---there are plenty of those. This is something different. It's the real story of what happens inside a man when he becomes a father. The identity crisis. The marriage strain. The suffocating weight of responsibility. The moments of grace that make it all worthwhile.

I write as a Christian, shaped by faith and trying to raise my children in that tradition. I also write as someone deeply influenced by Stoic philosophy---the ancient practice of focusing on what you can control, accepting what you cannot, and building character through discipline and reflection. These two streams run through everything I believe about fatherhood. They are not contradictory. They are complementary. The Stoics taught that virtue is its own reward; Christ taught that love is the greatest commandment. Both point to the same truth: who you are matters more than what you achieve.

This book is not perfect. Neither am I. But it is honest.

If you're holding your first child and wondering what you've gotten yourself into, this book is for you. If you're staring at a positive pregnancy test and feeling the first tremors of change, this book is for you. If you're three years in and still feel like you're faking it, this book is for you.

You are not alone. You are not failing. You are becoming.

Let's begin.

\vspace{2em}
\hfill---Tulio Soria
